```{r fig.height=3, fig.width=3, echo=FALSE, eval=FALSE}
# boxplot(mydata$bmi)
# boxplot(mydata$bmi, horizontal = TRUE, xlab = "BMI")
# title("BMI's of students")
```


---

# Convert sex into a factor level variable


```{r}
summary(mydata$sex)
mydata$sex <- as.factor(mydata$sex)
summary(mydata$sex)
```


---

<!-- class: inverse -->

# What other variables could we <br> convert to factor?
<!-- QQ: how to prevent title being cut off? -->

```{r}
mydata$smoked_ever <- as.factor(mydata$smoked_ever)
mydata$race4 <- as.factor(mydata$race4)
mydata$text_while_driving_30d <- as.factor(mydata$text_while_driving_30d)
mydata$grade <- as.factor(mydata$grade)
```

<!-- ## New summary of data set -->

```{r}
summary(mydata[,c("sex", "smoked_ever", "race4", "text_while_driving_30d", "grade")])
```


---

# Reordering factor levels

```{r}

```


--- 

# Access specific values in data set (2/2)

--- 
# Practice on your own

- create a folder on your computer
- set your working directory in RStudio to that folder
- make a new script and save it as `practice.R` in your folder

(Note: later using `Projects` will make this easier)


<!-- * In future workshops we will disucss the R workspace and working with R projects. -->
<!--     + The R workspace is a file where all the variables you created in your R session are saved  -->
<!--       - It is also specifies the directory where your work is being executed, which is important when loading and saving data sets -->
<!--     + Using an R project unifiies the workspace with code -->



# Variable types

* `num`numeric
* `int`: integer
* `fct`: factor
* `chr`: character
* `logi`: logical (TRUE, FALSE)



.pull-left[
* Entire column

```{r}
# Third column
mydata[, 3]
```
]

.pull-right[
* Entire row
```{r}
# Fourth row
mydata[4,]
```
]

