---
title: Practice Solutions to <br> Getting Started with R and RStudio 
author: Jessica Minnier, PhD & Meike Niederhausen, PhD
date: 'OCTRI Biostatistics, Epidemiology, Research & Design (BERD) Workshop <br><br> 2019/09/24'
output: 
  xaringan::moon_reader:
    css: [css/xaringan-themer-solns.css, css/my-theme.css]
    lib_dir: libs
    nature:
      highlightStyle: tomorrow #http://arm.rbind.io/slides/xaringan.html#77
      highlightLines: true
      highlightLanguage: r
      countIncrementalSlides: false
      titleSlideClass: ["left", "middle", "inverse"]
      ratio: "16:9"
    includes:
      in_header: ../header.html   
editor_options: 
  chunk_output_type: console
---

layout: true
  
<!-- <div class="my-footer"><span>bit.ly/berd_tidy</span></div>  -->

---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning=FALSE, 
  message=FALSE, 
  #fig.width=10.5, 
  #fig.height=4,
  fig.align = "center",
  rows.print=7,
  echo=TRUE,
  highlight = TRUE,
  prompt = FALSE, # IF TRUE adds a > before each code input
  comment = NA # PRINTS IN FRONT OF OUTPUT, default is '##' which comments out output
  #comment=NA
  )

# set ggplot theme
theme_set(theme_bw(base_size = 24))
```



```{r xaringan-themer, include = FALSE}
# creates xaringan theme
# devtools::install_github("gadenbuie/xaringanthemer")
library(xaringanthemer)
mono_light(
  base_color =  "#518c66", ## OHSU Marquam
  code_highlight_color = "#c0e8f5",
  link_color = "#38BDDE",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i","400i","700"),
  code_font_google   = NULL,
  text_font_size = "24px",
  code_font_size = "20px",
  header_h1_font_size = "45px",
  header_h2_font_size = "40px",
  header_h3_font_size = "35px",
  padding = "0em 2em 1em 2em",
  outfile = "css/xaringan-themer-solns.css"
)
```




# Practice questions 1

1. Open a new R script and type code/answers for next tasks in it. Save as `Practice1.R`

1. Create a vector of all integers from 4 to 10, and save it as `a1`. 

1. Create a vector of _even_ integers from 4 to 10, and save it as `a2`. 

1. What is the sum of `a1` and `a2`? 

1. What does the command `sum(a1)` do?

1. What does the command `length(a1)` do?

1. Use the commands to calculate the average of the values in `a1`.

1. The formula for the first $n$ integers is $n(n+1)/2$. Compute the sum of all integers from 1 to 100 to verify that this formula holds for $n=100$. 

1. Compute the sum of the squares of all integers from 1 to 100.

1. Take a break!

---

# Answers to practice questions 1

__#2__ Create a vector of all integers from 4 to 10, and save it as `a1`. 

__#3__ Create a vector of _even_ integers from 4 to 10, and save it as `a2`. 

```{r}
a1 <- 4:10
a2 <- c(4, 6, 8, 10)
# the following works as well:
a2 <- 2*(2:5)
# or
a2 <- seq(4, 10, by=2)
```

---

__#4__ What is the sum of `a1` and `a2`? 

````{r warning=TRUE}
a1+a2
````

Note that instead of giving an error, the terms of `a1` are repeated as needed since `a2` is longer than `a1`


---

__#5__ What does the command `sum(a1)` do?

```{r}
sum(a1)
```

`sum` adds up the values in the vector

<br>

__#6__ What does the command `length(a1)` do?

```{r}
length(a1)
```

`length` is the number of values in the vector

---

__#7__ Use the commands to calculate the average of the values in `a1`.

```{r}
sum(a1) / length(a1)

# this is equivalent
mean(a1)
```

---

__#8__ The formula for the first $n$ integers is $n(n+1)/2$. Compute the sum of all integers from 1 to 100 to verify that this formula holds for $n=100$. 

```{r}
sum(1:100)
# verify formula for n=100:
n=100
n * (n+1) / 2
```


---

__#9__ Compute the sum of the squares of all integers from 1 to 100.

```{r}
# The following code creates a vector of the squares of all integers from 1 to 100
(1:100)^2

# Now add the squares:
sum((1:100)^2)
```


---

```{r echo=FALSE}
mydata <- read.csv(here::here("01-getting-started-v2","data","yrbss_demo.csv"))
# mydata <- read.csv("data/yrbss_demo.csv")

# Add height column to data frame
mydata$height_m <- sqrt( mydata$weight_kg / mydata$bmi)

```


# Practice 2

1. Create a new script and save it as `Practice2.R`

1. Create data frames for males and females separately.

2. Do males and females have similar BMIs? Weights? Compares means, standard deviations, range, and boxplots.

3. Plot BMI vs. weight for each gender separately. Do they have similar relationships?

4. Are males or females more likely to be bullied in the past 12 months? Calculate the percentage bullied for each gender.

5. Are students that were bullied in the past year more likely to have smoked in the past? Does this vary by gender?


---

# Practice Answers (1/7)

__#2__ Create data frames for males and females separately.

```{r}
boys <- mydata[mydata$sex == "Male", ]
girls <- mydata[mydata$sex == "Female", ]
```

---

__#3__ Do males and females have similar BMIs? Weights? Compares means, standard deviations, range, and boxplots.

.pull-left-60[
```{r}
summary(boys$bmi); sd(boys$bmi)
summary(girls$bmi); sd(girls$bmi)
```

]
.pull-right-40[

```{r fig.height=5}
boxplot(mydata$bmi ~ mydata$sex)
```
]

---


__#4__ Plot BMI vs. weight for each gender separately. Do they have similar relationships?

.pull-left[
```{r fig.height=5, fig.width=4}
plot(boys$bmi, boys$weight)
```

]

.pull-right[

```{r fig.height=5, fig.width=4}
plot(girls$bmi, girls$weight)
```

]



---

__#5__ Are males or females more likely to be bullied in the past 12 months? Calculate the percentage bullied for each gender.

.pull-left[
```{r}
bullied_boys <- 
  boys[boys$bullied_past_12mo == TRUE,]
nrow(bullied_boys)
bullied_boys_prct <- 
  nrow(bullied_boys) / nrow(boys) * 100
bullied_boys_prct

# alternative
mean(boys$bullied_past_12mo, na.rm=TRUE)
```
]
.pull-right[
```{r}
# THIS DOESN'T WORK, need to use which(girls$bullied_past_12mo) but we didn't teach which()
bullied_girls <- 
  girls[girls$bullied_past_12mo == TRUE,]
nrow(bullied_girls)
bullied_girls_prct <- 
  nrow(bullied_girls) / nrow(girls) * 100
bullied_girls_prct

# alternative
mean(girls$bullied_past_12mo, na.rm=TRUE)
```
]

---


__#6__ Are students that were bullied in the past year more likely to have smoked in the past? Does this vary by gender?

```{r}
bullied_yes <- mydata[mydata$bullied_past_12mo == TRUE,]
bullied_no <- mydata[mydata$bullied_past_12mo == FALSE,]

# Not bullied students have higher proportion of smokers
summary(bullied_yes$smoked_ever)
summary(bullied_no$smoked_ever)
```
 
---

```{r}
## 6 ctd, Does it vary by gender? Not really
summary(bullied_yes[bullied_yes$sex == "Male", "smoked_ever"])
summary(bullied_yes[bullied_yes$sex == "Female", "smoked_ever"])

summary(bullied_no[bullied_no$sex == "Male", "smoked_ever"])
summary(bullied_no[bullied_no$sex == "Female", "smoked_ever"])
```


```{r, include=FALSE}
#knitr::purl("01-getting-started-v2/01_getting_started_Practice_Answers.Rmd")
pagedown::chrome_print(here::here("01-getting-started-v2","01_getting_started_Practice_Answers.html"))

```

