---
title: Getting Started with R and RStudio
author: Jessica Minnier, PhD & Meike Niederhausen, PhD
date: 'OCTRI Biostatistics, Epidemiology, Research & Design (BERD) Workshop <br><br> 2019/02/26 <br> Slides available at http://bit.ly/berd_r_intro <br> pdf version: http://bit.ly/berd_r_intro_pdf'
output: 
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, mycss.css]
    lib_dir: libs
    nature:
      highlightStyle: rainbow #http://arm.rbind.io/slides/xaringan.html#77
      highlightLines: true
      highlightLanguage: r
      countIncrementalSlides: false
    includes:
      in_header: ../header.html   
    #css: ../jminnier.css
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

library(tidyverse)

knitr::opts_chunk$set(
  warning=FALSE, 
  message=FALSE, 
  fig.width=10.5, 
  fig.height=4,
  fig.align = "center",
  rows.print=16,
  echo=TRUE,
  highlight = TRUE,
  prompt = TRUE, # IF TRUE adds a > before each code input
  comment=NA # PRINTS IN FRONT OF OUTPUT, default is '##' which comments out output
  #comment=NA
  )

theme_set(theme_bw(base_size = 24))
```


---

# Practice questions

1. Create a vector of all integers from 4 to 10, and save it as `a1`. 

2. Create a vector of _even_ integers from 4 to 10, and save it as `a2`. 

3. What is the sum of `a1` and `a2`? 

4. What does the command `sum(a1)` do?

5. What does the command `length(a1)` do?

6. Use the commands to calculate the average of the values in `a1`.

7. The formula for the first $n$ integers is $n(n+1)/2$. Compute the sum of all integers from 1 to 100 to verify that this formula holds for $n=100$. 

8. Compute the sum of the squares of all integers from 1 to 100.

9. Take a break!


---

# Answers to practice questions (1/4)

__#1__ Create a vector of all integers from 4 to 10, and save it as `a1`. 

__#2__ Create a vector of _even_ integers from 4 to 10, and save it as `a2`. 

```{r}
a1 <- 4:10
a2 <- c(4, 6, 8, 10)
```


__#3__ What is the sum of `a1` and `a2`? 

````{r warning=TRUE}
a1+a2
````

Note that instead of giving an error, the terms of `a1` are repeated as needed since `a2` is longer than `a1`


---

# Answers to practice questions (2/4)

__#4__ What does the command `sum(a1)` do?

```{r}
sum(a1)
```

`sum` adds up the values in the vector

<br>

__#5__ What does the command `length(a1)` do?

```{r}
length(a1)
```

`length` is the number of values in the vector

---

# Answers to practice questions (3/4)

__#6__ Use the commands to calculate the average of the values in `a1`.

```{r}
sum(a1) / length(a1)
```


__#7__ The formula for the first $n$ integers is $n(n+1)/2$. Compute the sum of all integers from 1 to 100 to verify that this formula holds for $n=100$. 

```{r}
sum(1:100)
# verify formula for n=100:
n=100
n * (n+1) / 2
```


---

# Answers to practice questions (4/4)

__#8__ Compute the sum of the squares of all integers from 1 to 100.

```{r}
# The following code creates a vector of the squares of all integers from 1 to 100
(1:100)^2

# Now add the squares:
sum((1:100)^2)
```


---

```{r echo=FALSE}
mydata <- read.csv(url("http://bit.ly/berd_data_csv"))

# Add height column to data frame
mydata$height_m <- sqrt( mydata$weight_kg / mydata$bmi)

```


# Practice

1. Create data frames for males and females separately.

2. Do males and females have similar BMI's? Weights? Compares means, standard deviations, range, and boxplots.

3. Plot BMI vs. weight for each gender separately. Do they have similar relationships?

4. Are males or females more likely to be bullied in the past 12 months? Calculate the percentage bullied for each gender.

5. Are students that were bullied in the past year more likely to have smoked in the past? Does this vary by gender?


---

# Practice Answers (1/7)

__#1__ Create data frames for males and females separately.

```{r}
boys <- mydata[mydata$sex == "Male", ]
girls <- mydata[mydata$sex == "Female", ]
```


---

# Practice Answers (2/7)

__#2__ Do males and females have similar BMI's? Weights? Compares means, standard deviations, range, and boxplots.

```{r}
summary(boys$bmi); sd(boys$bmi)
summary(girls$bmi); sd(girls$bmi)
```

---

# Practice Answers (3/7)

__#2__ cont'd

```{r fig.height=5}
boxplot(mydata$bmi ~ mydata$sex)
```


---

# Practice Answers (4/7)

__#3__ Plot BMI vs. weight for each gender separately. Do they have similar relationships?

.pull-left[
```{r fig.height=5, fig.width=4}
plot(boys$bmi, boys$weight)
```

]

.pull-right[

```{r fig.height=5, fig.width=4}
plot(girls$bmi, girls$weight)
```

]



---

# Practice Answers (5/7)

__#4__ Are males or females more likely to be bullied in the past 12 months? Calculate the percentage bullied for each gender.

```{r}
bullied_boys <- boys[boys$bullied_past_12mo == TRUE,]
nrow(bullied_boys)
bullied_boys_prct <- nrow(bullied_boys) / nrow(boys) * 100; bullied_boys_prct

bullied_girls <- girls[girls$bullied_past_12mo == TRUE,]
nrow(bullied_girls)
bullied_girls_prct <- nrow(bullied_girls) / nrow(girls) * 100; bullied_girls_prct
```


---

# Practice Answers (6/7)

__#5__ Are students that were bullied in the past year more likely to have smoked in the past? Does this vary by gender?

```{r}
bullied_yes <- mydata[mydata$bullied_past_12mo == TRUE,]
bullied_no <- mydata[mydata$bullied_past_12mo == FALSE,]

# Not bullied students have higher proportion of smokers
summary(bullied_yes$smoked_ever)
summary(bullied_no$smoked_ever)
```
 
---

# Practice Answers (7/7)

__#5__ cont'd

```{r}
# Vary by gender? Not really.
summary(bullied_yes[bullied_yes$sex == "Male", "smoked_ever"])
summary(bullied_yes[bullied_yes$sex == "Female", "smoked_ever"])

summary(bullied_no[bullied_no$sex == "Male", "smoked_ever"])
summary(bullied_no[bullied_no$sex == "Female", "smoked_ever"])
```


```{r, include=FALSE}
#knitr::purl("01-getting-started/01_getting_started_Practice_Answers.Rmd")
```

