---
title: "Advanced Wrangling Examples"
author: "Jessica Minnier & Meike Niederhausen"
date: '`r Sys.Date()`'
output:   
  html_document:
    highlight: pygments
    keep_md: no
    number_sections: yes
    smart: yes
    theme: yeti
    toc: yes
    toc_float: yes
---


```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(skimr)
library(janitor)
library(jmmisc)
#### Global chunk options -----------------------------

knitr::opts_chunk$set(
  eval       = TRUE,    # whether to run code in code chunk
  include    = TRUE,    # whether to include the chunk output
  echo       = TRUE,   # Whether to show code chunk in final output
  error      = TRUE,    # whether to display error messages
  message    = FALSE,   # whether to preserve messages
  warning    = FALSE,   # whether to preserve warnings
  comment    = "#>",    # a character string to append at start
                        # of each line of results in final document
  tidy       = FALSE,   # whether to tidy code chunks for display
  dpi        = 96, 
  fig.width  = 6,       # consistent width for figures
  fig.asp    = 0.618,   # the golden ratio, can be adjusted in individual chunks
  out.width  = "100%",   # controls the output size
  fig.align  = "center" # give plot room to breathe
)
```


# Import data

```{r}
# install.packages("tidyverse","janitor")  
library(tidyverse)
library(lubridate)  
library(janitor)
demo_data <- read_csv("data/yrbss_demo.csv")
qn_data <- read_csv("data/yrbss_qn.csv")
```

# Look at the data

```{r}
glimpse(demo_data)
glimpse(qn_data)
```



# Join the data

Make sure all records are unique. How many overlap?
```{r}
length(unique(demo_data$record))

length(unique(qn_data$record))

length(intersect(demo_data$record, qn_data$record))
```

Looks like all records in qn_data are also in demo_data. Is that right?

```{r}
table(qn_data$record%in%demo_data$record)
```


```{r}
all_data <- left_join(demo_data, qn_data)
glimpse(all_data)
```

```{r}
all_data_clean <- all_data %>%
  # numeric grade
  mutate(grade_num = as.numeric(str_replace(grade,"th","")),
         female = 1*(sex=="Female"))%>%
  # numeric age (<= 12 to 12 and >= 18 to 18)
  separate(age,c("age_num"),"year",remove = FALSE) %>%
  mutate(age_num = as.numeric(age_num)) %>%
  # add bmi categories
  mutate(
    bmi_cat = case_when(
      bmi < 18.5 ~ "underweight",               # condition ~ new_value
      bmi >= 18.5 & bmi <= 24.9 ~ "normal",
      bmi > 24.9 & bmi <= 29.9 ~ "overweight",
      bmi > 29.9 ~ "obese")
    )

glimpse(all_data_clean)
```

Check that everything worked the right way:

```{r}
all_data_clean %>% tabyl(race4, race7)
all_data_clean %>% tabyl(sex, female)
all_data_clean %>% tabyl(age, age_num)
all_data_clean %>% group_by(bmi_cat) %>% summarize(min(bmi),max(bmi))
```

Let's tackle the question answers. First rename them to something meaninful and then look at the levels.

```{r}
all_data_clean <- all_data_clean %>%
  # q8 = How often wear bicycle helmet
  # q12 = Texted while driving
  # q31 = Ever smoked
  # qn24 = Bullied past 12 months
  rename(q_bike_helmet = q8,
         q_text_while_driving = q12,
         q_smoked_ever = q31,
         q_bullied_past_12mo = qn24
         )

all_data_clean %>% tabyl(q_bike_helmet)
all_data_clean %>% tabyl(q_text_while_driving)
all_data_clean %>% tabyl(q_smoked_ever)
all_data_clean %>% tabyl(q_bullied_past_12mo)
```

Convert the true/false and yes/no to numeric.

```{r}
all_data_clean <- all_data_clean %>%
  mutate(q_smoked_ever = 1*(q_smoked_ever=="Yes"),
         q_bullied_past_12mo = 1*(q_bullied_past_12mo))
```

# Missingness

How many rows do we have?

```{r}
all_data_clean %>% nrow
```

How many rows are left if we take out any row with missingness?

```{r}
all_data_clean %>% na.omit %>% nrow
```

Could replace missing `sex` with `Unknown`:

```{r}
newdata <- all_data_clean %>%
  mutate(sex2 = replace_na(sex, "Unknown"))

newdata %>% tabyl(sex, sex2)
```

Could replace all NAs in the question columns with another string, i.e. "No answer":

```{r}
all_data_clean %>% 
  mutate_at(vars(contains("q")),list(~replace_na(.,"No answer"))) %>%
  tabyl(q_text_while_driving)
```


Could remove all data with missingness in sex and age

```{r}
newdata <- all_data_clean %>% drop_na(sex, age)
nrow(newdata)
```


# Summarize

What proportion were bullied in the last 12 months, by sex? Include counts of each sex category.

```{r}
all_data_clean %>% 
  group_by(sex) %>% 
  summarize(
    "n"=n(),
    "proportion_bullied"=round(mean(q_bullied_past_12mo,na.rm=T),3))
```

