---
title: Data Wrangling in R with the Tidyverse (Part 2) - Practice Solutions
author: "Jessica Minnier, PhD & Meike Niederhausen, PhD<br><span style = 'font-size: 80%;'>[OCTRI Biostatistics, Epidemiology, Research & Design (BERD) Workshop](https://www.ohsu.edu/xd/research/centers-institutes/octri/education-training/octri-research-forum.cfm) </span>"
date: "<span style = 'font-size: 80%;'>2019/04/25 (Part 2)<br><br><br> `r icon::fa('link')` slides: [bit.ly/berd_tidy1](http://bit.ly/berd_tidy1) <br> `r icon::fa('file-pdf')` pdf:  [bit.ly/berd_tidy1_pdf](http://bit.ly/berd_tidy1_pdf)</span>"
output: 
  xaringan::moon_reader:
    css: [css/xaringan-themer-solns.css, css/my-theme.css]
    lib_dir: libs
    nature:
      highlightStyle: tomorrow #http://arm.rbind.io/slides/xaringan.html#77
      highlightLines: true
      highlightLanguage: r
      countIncrementalSlides: false
      titleSlideClass: ["left", "middle", "inverse"]
      ratio: "16:9"
    includes:
      in_header: ../header.html   
editor_options: 
  chunk_output_type: console
---

layout: true
  
<!-- <div class="my-footer"><span>bit.ly/berd_tidy</span></div>  -->

---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning=FALSE, 
  message=FALSE, 
  #fig.width=10.5, 
  #fig.height=4,
  fig.align = "center",
  rows.print=7,
  echo=TRUE,
  highlight = TRUE,
  prompt = FALSE, # IF TRUE adds a > before each code input
  comment = NA # PRINTS IN FRONT OF OUTPUT, default is '##' which comments out output
  #comment=NA
  )

# set ggplot theme
theme_set(theme_bw(base_size = 24))
```



```{r xaringan-themer, include = FALSE}
# creates xaringan theme
# devtools::install_github("gadenbuie/xaringanthemer")
library(xaringanthemer)
mono_light(
  base_color =  "#518c66", ## OHSU Marquam
  code_highlight_color = "#c0e8f5",
  link_color = "#38BDDE",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i","400i","700"),
  code_font_google   = NULL,
  text_font_size = "24px",
  code_font_size = "20px",
  header_h1_font_size = "45px",
  header_h2_font_size = "40px",
  header_h3_font_size = "35px",
  padding = "0em 2em 1em 2em",
  outfile = "css/xaringan-themer-solns.css"
)
```

# Load the data and packages

```{r}
# install.packages("tidyverse","janitor","glue")  
library(tidyverse)
library(lubridate)  
library(janitor)
library(glue)
demo_data <- read_csv("data/yrbss_demo.csv")
qn_data <- read_csv("data/yrbss_qn.csv")
```

---

# Practice 1

1. Add a column of `1`'s to `qn_data` called `qn_yes` and save the resulting data as `qn_data2`.

1. Join `demo_data` and `qn_data2` by column `record`. Keep all rows from `demo_data` and only rows from `qn_data2` that match records in `demo_data`. Call the resulting data `all_data`.

1. Create a `tabyl()` of `qn_yes` for the data `all_data`.

1. Create a 2x2 table of `qn_yes` vs `grade`.

Note about the data:

- q8 = How often wear bicycle helmet
- q12 = Texted while driving
- q31 = Ever smoked
- qn24 = Bullied past 12 months

---

```{r,}
qn_data2 <- qn_data %>% add_column(qn_yes = 1)
all_data <- left_join(demo_data, qn_data2)
all_data %>% tabyl(qn_yes)
all_data %>% tabyl(qn_yes,grade)
```

---

# Practice 2


1. Make `DBP_wide` into a long dataframe based on the repeated DBP columns and save it as `DBP_long`.
1. Clean up the visit column of `DBP_long` so that the values are 1, 2, 3, and save it as `DBP_long`.

1. Make `DBP_long` wide with column names `visit.1, visit.2, visit.3` for the DBP values, and save it as `DBP_wide2`.

1. Join `DBP_long` with `BP_long2` so that we have one data frame with columns id, sex, visit, SBP, DBP, and age. Save this as `BP_both_long`.

---

# Practice 2 Initial Data

Copy and paste the code below into R to create the datasets:

```{r}
DBP_wide <- tibble(id = letters[1:4],
                  sex = c("F", "M", "M", "F"),
                  v1.DBP = c(88, 84, 102, 70),
                  v2.DBP = c(78, 78, 96, 76),
                  v3.DBP = c(94, 82, 94, 74),
                  age=c(23, 56, 41, 38)
                  )
BP_wide <- tibble(id = letters[1:4],
                     sex = c("F", "M", "M", "F"),
                     SBP_v1 = c(130, 120, 130, 119),
                     SBP_v2 = c(110, 116, 136, 106),
                     SBP_v3 = c(112, 122, 138, 118))
BP_long <- BP_wide %>% 
  gather(key = "visit", value = "SBP", SBP_v1:SBP_v3)
BP_long2 <- BP_long %>% 
  mutate(visit = str_replace(visit,"SBP_v","")) 
```
---

# Practice 2 solutions (1/2)

.pull-left[
```{r}
DBP_long <- DBP_wide %>%
  gather(key = "visit", value = "DBP", 
         v1.DBP, v2.DBP, v3.DBP) %>%
  mutate(visit = str_replace(
    visit, c("v"), "")) %>%
  mutate(visit = str_replace(
    visit, ".DBP", ""))  
DBP_long
```
]
.pull-right[
```{r}
DBP_wide2 <- DBP_long %>% 
  spread(
    key = "visit", value = "DBP",
    sep=".") # specify separating character
DBP_wide2
```
]
---

# Practice 2 solutions (2/2)

```{r}
BP_both_long <- left_join(BP_long2, DBP_long, by = c("id", "sex", "visit"))
BP_both_long
```

---

# Practice 3

```{r}
messy_data <- tibble(NAME = c("J N","A C","D E"), 
                     `months follow up` = c("", 10, 11), 
                     `Date of visit` = c("July 31, 2003", "Nov 12, 2005", "Aug 3, 2007"))
```

1. Clean column names with `clean_names()`.
1. Replace missing ("") data in `months_follow_up` with NA.
1. Convert `months_follow_up` to a numeric variable.
1. Convert `date_of_visit` to a date.
1. Create a column called `date_last_visit` that is the date of visit *plus* months of follow up.
1. Remove rows (cases) with missing data in `months_follow_up`.
1. Remove the spaces in `name`.

```{r}
messy_data
```


---

# Practice solutions 3 (1/2)

```{r}
clean_data <- messy_data %>% 
  clean_names() %>%
  mutate(
    months_follow_up = replace_na(months_follow_up,""),
    months_follow_up = as.numeric(months_follow_up),
    date_of_visit = mdy(date_of_visit),
    date_last_visit = date_of_visit + months(months_follow_up))
clean_data
```

---

# Practice solutions 3 (2/2)

```{r}

clean_data <- clean_data %>%
  drop_na(months_follow_up) %>%
  mutate(name = str_replace_all(name," ",""))
clean_data
```

