---
title: Data Wrangling in R with the Tidyverse (Part 1) - Practice Solutions
author: "Jessica Minnier, PhD & Meike Niederhausen, PhD<br><span style = 'font-size: 80%;'>[OCTRI Biostatistics, Epidemiology, Research & Design (BERD) Workshop](https://www.ohsu.edu/xd/research/centers-institutes/octri/education-training/octri-research-forum.cfm) </span>"
date: "<span style = 'font-size: 80%;'>2019/04/18 (Part 1)<br><em>and again!</em> 2019/05/16 (Part 1)<br><br><br> `r icon::fa('link')` slides: [bit.ly/berd_tidy1](http://bit.ly/berd_tidy1) <br> `r icon::fa('file-pdf')` pdf:  [bit.ly/berd_tidy1_pdf](http://bit.ly/berd_tidy1_pdf)</span>"
output: 
  xaringan::moon_reader:
    css: [css/xaringan-themer-solns.css, css/my-theme.css]
    lib_dir: libs
    nature:
      highlightStyle: tomorrow #http://arm.rbind.io/slides/xaringan.html#77
      highlightLines: true
      highlightLanguage: r
      countIncrementalSlides: false
      titleSlideClass: ["left", "middle", "inverse"]
      ratio: "16:9"
    includes:
      in_header: ../header.html   
editor_options: 
  chunk_output_type: console
---

layout: true
  
<!-- <div class="my-footer"><span>bit.ly/berd_tidy</span></div>  -->

---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning=FALSE, 
  message=FALSE, 
  #fig.width=10.5, 
  #fig.height=4,
  fig.align = "center",
  rows.print=7,
  echo=TRUE,
  highlight = TRUE,
  prompt = FALSE, # IF TRUE adds a > before each code input
  comment = NA # PRINTS IN FRONT OF OUTPUT, default is '##' which comments out output
  #comment=NA
  )

# set ggplot theme
theme_set(theme_bw(base_size = 24))
```



```{r xaringan-themer, include = FALSE}
# creates xaringan theme
# devtools::install_github("gadenbuie/xaringanthemer")
library(xaringanthemer)
mono_light(
  base_color =  "#518c66", ## OHSU Marquam
  code_highlight_color = "#c0e8f5",
  link_color = "#38BDDE",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i","400i","700"),
  code_font_google   = NULL,
  text_font_size = "24px",
  code_font_size = "20px",
  header_h1_font_size = "45px",
  header_h2_font_size = "40px",
  header_h3_font_size = "35px",
  padding = "0em 2em 1em 2em",
  outfile = "css/xaringan-themer-solns.css"
)
```

# Load the data and packages

```{r}
# install.packages("tidyverse")
library(tidyverse)  
library(lubridate)
demo_data <- read_csv("data/yrbss_demo.csv")
```

---

# Practice 1

1. Import `demo_data.csv` in the `data` folder if you haven't already done so.

1. Filter `newdata` to only keep Asian or Native Hawaiian/other PI subjects that are in the 9th grade, and save again as `newdata`.

1. Filter `newdata` to remove subjects younger than 13, and save as `newdata`.

1. Remove the column `race4`, and save as `newdata`.

1. How many rows does the resulting `newdata` have? How many columns?

---

# Practice 1 Solutions (1/2)

```{r}
newdata <- demo_data %>%
  filter(race7 %in% c("Asian","Native Hawaiian/other PI"), 
         grade == "9th",
         age != "12 years old or younger") %>%
  select(-race4)
newdata
```

---

# Practice 1 Solutions (2/2)

```{r}
dim(newdata) # both nrow and ncol
nrow(newdata)
ncol(newdata)
```

---

# Practice 2

Do the following data wrangling steps in order so that the output from the previous step is the input for the next step.
Save the results in each step as `newdata`.

1. Import `demo_data.csv` in the `data` folder if you haven't already done so.

1. Create a variable called `grade_num` that has the numeric grade number (use `as.numeric`).

1. Filter the data to keep only students in grade 11 or higher.

1. Filter out rows when `bmi` is `NA`.

1. Create a binary variable called `bmi_normal` that is equal to 1 when `bmi` is between 18.5 to 24.9 and 0 when it is outside that range.

1. Arrange by `grade_num` from highest to lowest

1. Save all output to `newdata`.

---

# Practice 2 Solutions (1/2)

```{r}
newdata <- demo_data %>%
  separate(grade, c("grade_num"), sep = "th") %>%
  mutate(grade_num = as.numeric(grade_num)) %>%
  filter(grade_num >= 11,
         !is.na(bmi)) %>%
  mutate(
    bmi_normal = case_when(
      (18.5 <= bmi) & (bmi <= 24.9) ~ 1,
      bmi > 24.9 ~ 0,
      bmi < 18.5 ~ 0,
      )
    ) %>%
  arrange(desc(grade_num))
newdata
```

---

# Practice 2 Solutions - Alternative (2/2)

```{r}
newdata <- demo_data %>%
  mutate(
    grade_num = str_replace(grade,"th",""),
    grade_num = as.numeric(grade_num),
    ) %>%
  filter(grade_num >= 11,
         !is.na(bmi)) %>%
  mutate(
    bmi_normal = case_when(
      (18.5 <= bmi) & (bmi <= 24.9) ~ 1,
      TRUE ~ 0, # TRUE is like "else", also changes NAs if any
      )
    ) %>%
  arrange(desc(grade_num))
newdata
```

