---
title: "Gapminder data prep"
author: "meike"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: cerulean
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height=3, fig.width=4, message = F)
```

```{r}
library(readr)
library(tidyverse)
library(janitor)
library(skimr)
library(knitr)

```

# Datasets

R name  | file name |  year subsetted
------------|--------|--------
LifeExpectancyYrs | life_expectancy_years.csv | 2011
FemaleLiteracyRate | literacy_rate_adult_female_percent_of_females_ages_15_above.csv | 2011
WaterSourcePrct | at_least_basic_water_source_overall_access_percent.csv | 2011
CO2emissions | raw_data/co2_emissions_tonnes_per_person.csv
ElectricityUsePP | raw_data/electricity_use_per_person.csv
FoodSupplykcPPD | raw_data/food_supply_kilocalories_per_person_and_day.csv
IncomePP | raw_data/income_per_person_gdppercapita_ppp_inflation_adjusted.csv | up to 2040
population | raw_data/population_total.csv  | up to 2100



# Life expectancy

* Complete data for 1970-2016

```{r}
LifeExpectancyYrs_temp <- read_csv("raw_data/life_expectancy_years.csv")
LifeExpectancyYrs <- clean_names(LifeExpectancyYrs_temp)
# glimpse(LifeExpectancyYrs)
dim(LifeExpectancyYrs)
# names(LifeExpectancyYrs)
```


## skim

```{r}
LifeExpectancyYrs_skim <- skim(LifeExpectancyYrs)
LifeExpectancyYrs_skim
```

## years with fewest missing data

```{r}
LifeExpectancyYrs_minNA <- LifeExpectancyYrs_skim %>% 
  select(year = skim_variable, n_miss_LifeExpectancyYrs = n_missing) %>% 
  filter(year != "country") %>% 
  arrange(n_miss_LifeExpectancyYrs, desc(year)) 
head(LifeExpectancyYrs_minNA, 15)
```


```{r eval=FALSE, echo=FALSE}
str(LifeExpectancyYrs_skim)
names(LifeExpectancyYrs_skim)
```




## Restrict to 2011

* 2011 is year with most data for Adult female FemaleLiteracyRate rate

```{r}
LifeExpectancyYrs_11 <- LifeExpectancyYrs %>% select(country, "x2011") %>% 
  rename(LifeExpectancyYrs = x2011)
LifeExpectancyYrs_11
```



# Literacy: Adult female FemaleLiteracyRate rate

```{r}
FemaleLiteracyRate_temp <- read_csv("raw_data/literacy_rate_adult_female_percent_of_females_ages_15_above.csv")
FemaleLiteracyRate <- clean_names(FemaleLiteracyRate_temp)
# glimpse(FemaleLiteracyRate)
dim(FemaleLiteracyRate)
# names(FemaleLiteracyRate)
```


## skim

```{r}
FemaleLiteracyRate_skim <- skim(FemaleLiteracyRate)
FemaleLiteracyRate_skim
```

## years with fewest missing data

```{r}
FemaleLiteracyRate_minNA <- FemaleLiteracyRate_skim %>% 
  select(year = skim_variable, n_miss_FemaleLiteracyRate = n_missing) %>% 
  filter(year != "country") %>% 
  arrange(n_miss_FemaleLiteracyRate, desc(year)) 
head(FemaleLiteracyRate_minNA, 15)
```


```{r eval=FALSE, echo=FALSE}
str(FemaleLiteracyRate_skim)
names(FemaleLiteracyRate_skim)
```


## Restrict to 2011

* 2011 is year with most data for FemaleLiteracyRate rate

```{r}
FemaleLiteracyRate_11 <- FemaleLiteracyRate %>% select(country, "x2011") %>% 
  rename(FemaleLiteracyRate = x2011)
FemaleLiteracyRate_11
```



# Water access

* n =194
* Complete data for 2011-2012
* distribution skewed right

```{r}
WaterSourcePrct_temp <- read_csv("raw_data/at_least_basic_water_source_overall_access_percent.csv")
WaterSourcePrct <- clean_names(WaterSourcePrct_temp)
# glimpse(WaterSourcePrct)
dim(WaterSourcePrct)
# names(WaterSourcePrct)
```


## skim

```{r}
WaterSourcePrct_skim <- skim(WaterSourcePrct)
WaterSourcePrct_skim
```

## years with fewest missing data

```{r}
WaterSourcePrct_minNA <- WaterSourcePrct_skim %>% 
  select(year = skim_variable, n_miss_WaterSourcePrct = n_missing) %>% 
  filter(year != "country") %>% 
  arrange(n_miss_WaterSourcePrct, desc(year)) 
head(WaterSourcePrct_minNA, 15)
```


```{r eval=FALSE, echo=FALSE}
str(WaterSourcePrct_skim)
names(WaterSourcePrct_skim)
```



## Restrict to 2011

* 2011 is year with most data for Adult female FemaleLiteracyRate rate

```{r}
WaterSourcePrct_11 <- WaterSourcePrct %>% select(country, "x2011") %>% 
  rename(WaterSourcePrct = x2011)
WaterSourcePrct_11

skim(WaterSourcePrct_11)

# WaterSourcePrct_11 %>% tabyl(WaterSourcePrct)
```


Split data into quartiles:

* group 1: [0, 74.9%]
* group 2: (74.9%, 93.5%]
* group 2: (93.5%, 99.1%]
* group 2: (99.1%, 100%]

### viz

```{r}
ggplot(WaterSourcePrct_11, aes(x=WaterSourcePrct)) +
  geom_histogram()

ggplot(WaterSourcePrct_11, aes(x=WaterSourcePrct)) +
  geom_bar()

ggplot(WaterSourcePrct_11, aes(y=WaterSourcePrct)) +
  geom_boxplot()
```

### Categorical variable for quartiles


```{r}
WaterSourcePrct_11 <- WaterSourcePrct_11 %>% 
  mutate(
    WaterSourcePrct_2011_quart = case_when(
      WaterSourcePrct <= 74.9 ~ "Q1",
      WaterSourcePrct > 74.9 & WaterSourcePrct <= 93.5 ~ "Q2",
      WaterSourcePrct > 93.5 & WaterSourcePrct <= 99.1 ~ "Q3",
      WaterSourcePrct > 99.1 ~ "Q4"
    ))

WaterSourcePrct_11 %>% tabyl(WaterSourcePrct_2011_quart)
```

# CO2 emissions


```{r}
CO2emissions_temp <- read_csv("raw_data/co2_emissions_tonnes_per_person.csv")
CO2emissions <- clean_names(CO2emissions_temp)
# glimpse(CO2emissions)
dim(CO2emissions)
# names(CO2emissions)
```


## skim

```{r}
CO2emissions_skim <- skim(CO2emissions)
CO2emissions_skim
```

## years with fewest missing data

```{r}
CO2emissions_minNA <- CO2emissions_skim %>% 
  select(year = skim_variable, n_miss_CO2emissions = n_missing) %>% 
  filter(year != "country") %>% 
  arrange(n_miss_CO2emissions, desc(year)) 
head(CO2emissions_minNA, 15)
```


```{r eval=FALSE, echo=FALSE}
str(CO2emissions_skim)
names(CO2emissions_skim)
```

## Restrict to 2011

```{r}
CO2emissions_11 <- CO2emissions %>% select(country, "x2011") %>% 
  rename(CO2emissions = x2011)
CO2emissions_11

skim(CO2emissions_11)
```



# Electricity use


```{r}
ElectricityUsePP_temp <- read_csv("raw_data/electricity_use_per_person.csv")
ElectricityUsePP <- clean_names(ElectricityUsePP_temp)
# glimpse(ElectricityUsePP)
dim(ElectricityUsePP)
# names(ElectricityUsePP)
```


## skim

```{r}
ElectricityUsePP_skim <- skim(ElectricityUsePP)
ElectricityUsePP_skim
```

## years with fewest missing data

```{r}
ElectricityUsePP_minNA <- ElectricityUsePP_skim %>% 
  select(year = skim_variable, n_miss_ElectricityUsePP = n_missing) %>% 
  filter(year != "country") %>% 
  arrange(n_miss_ElectricityUsePP, desc(year)) 
head(ElectricityUsePP_minNA, 15)
```


```{r eval=FALSE, echo=FALSE}
str(ElectricityUsePP_skim)
names(ElectricityUsePP_skim)
```

## Restrict to 2011

```{r}
ElectricityUsePP_11 <- ElectricityUsePP %>% select(country, "x2011") %>% 
  rename(ElectricityUsePP = x2011)
ElectricityUsePP_11

skim(ElectricityUsePP_11)
```
# Food supply

food_supply_kilocalories_per_person_and_day


```{r}
FoodSupplykcPPD_temp <- read_csv("raw_data/food_supply_kilocalories_per_person_and_day.csv")
FoodSupplykcPPD <- clean_names(FoodSupplykcPPD_temp)
# glimpse(FoodSupplykcPPD)
dim(FoodSupplykcPPD)
# names(FoodSupplykcPPD)
```


## skim

```{r}
FoodSupplykcPPD_skim <- skim(FoodSupplykcPPD)
FoodSupplykcPPD_skim
```

## years with fewest missing data

```{r}
FoodSupplykcPPD_minNA <- FoodSupplykcPPD_skim %>% 
  select(year = skim_variable, n_miss_FoodSupplykcPPD = n_missing) %>% 
  filter(year != "country") %>% 
  arrange(n_miss_FoodSupplykcPPD, desc(year)) 
head(FoodSupplykcPPD_minNA, 15)
```


```{r eval=FALSE, echo=FALSE}
str(FoodSupplykcPPD_skim)
names(FoodSupplykcPPD_skim)
```

## Restrict to 2011

```{r}
FoodSupplykcPPD_11 <- FoodSupplykcPPD %>% select(country, "x2011") %>% 
  rename(FoodSupplykcPPD = x2011)
FoodSupplykcPPD_11

skim(FoodSupplykcPPD_11)
```


# Income

income_per_person_gdppercapita_ppp_inflation_adjusted


```{r}
IncomePP_temp <- read_csv("raw_data/income_per_person_gdppercapita_ppp_inflation_adjusted.csv")
IncomePP_temp2 <- clean_names(IncomePP_temp)
# glimpse(IncomePP)
dim(IncomePP_temp2)
# names(IncomePP_temp2)
tail(names(IncomePP_temp2))

IncomePP <- IncomePP_temp2 %>% 
  select(country:x2019)
dim(IncomePP)
```


## skim

```{r}
IncomePP_skim <- skim(IncomePP)
IncomePP_skim
```

## years with fewest missing data

```{r}
IncomePP_minNA <- IncomePP_skim %>% 
  select(year = skim_variable, n_miss_IncomePP = n_missing) %>% 
  filter(year != "country") %>% 
  arrange(n_miss_IncomePP, desc(year)) 
head(IncomePP_minNA, 15)
```


```{r eval=FALSE, echo=FALSE}
str(IncomePP_skim)
names(IncomePP_skim)
```

## Restrict to 2011

```{r}
IncomePP_11 <- IncomePP %>% select(country, "x2011") %>% 
  rename(IncomePP = x2011)
IncomePP_11

skim(IncomePP_11)
```

# Population

population_total.csv


```{r}
population_temp <- read_csv("raw_data/population_total.csv")
population_temp2 <- clean_names(population_temp)
# glimpse(population)
dim(population_temp2)
# names(population_temp2)
tail(names(population_temp2))

population <- population_temp2 %>% 
  select(country:x2019)
dim(population)
```


## skim

```{r}
population_skim <- skim(population)
population_skim
```

## years with fewest missing data

```{r}
population_minNA <- population_skim %>% 
  select(year = skim_variable, n_miss_population = n_missing) %>% 
  filter(year != "country") %>% 
  arrange(n_miss_population, desc(year)) 
head(population_minNA, 15)
```


```{r eval=FALSE, echo=FALSE}
str(population_skim)
names(population_skim)
```


## Restrict to 2011

```{r}
population_11 <- population %>% select(country, "x2011") %>% 
  rename(population = x2011)
population_11

skim(population_11)
```

# Compare years with fewest missing values

Looks like 2011 is the most recent year with the fewest missing values


```{r}
as.data.frame(ls()) %>% filter(str_detect(`ls()`, "_minNA"))
```


```{r}
minNA_join <- Reduce(full_join,list(
  CO2emissions_minNA, ElectricityUsePP_minNA, FoodSupplykcPPD_minNA, IncomePP_minNA, LifeExpectancyYrs_minNA, FemaleLiteracyRate_minNA, population_minNA, WaterSourcePrct_minNA
  ))

dim(minNA_join) # lots of missing values
minNA_join # lots of missing values

minNA_join_noNArow <- minNA_join %>% drop_na()
minNA_join_noNArow %>% kable()
```



# Joins

## Life expectancy & Adult female FemaleLiteracyRate rate

```{r}
LifeExpectancyYrs_FemaleLiteracyRate_2011 <- full_join(LifeExpectancyYrs_11, FemaleLiteracyRate_11,
                                 by = c("country")
                                 )
LifeExpectancyYrs_FemaleLiteracyRate_2011


skim(LifeExpectancyYrs_FemaleLiteracyRate_2011)
```


### save joined dataset

```{r}
save(LifeExpectancyYrs_FemaleLiteracyRate_2011, file = "LifeExpectancyYrs_FemaleLiteracyRate_2011.RData")
write_csv(LifeExpectancyYrs_FemaleLiteracyRate_2011, path = "LifeExpectancyYrs_FemaleLiteracyRate_2011.csv")
```

## Add WaterSourcePrct access 2011 to `LifeExpectancyYrs_FemaleLiteracyRate_2011`

```{r}
dim(LifeExpectancyYrs_FemaleLiteracyRate_2011)
dim(WaterSourcePrct_11)

LifeExpectancyYrs_FemaleLiteracyRate_WaterSourcePrct_2011 <- full_join(LifeExpectancyYrs_FemaleLiteracyRate_2011, WaterSourcePrct_11,
                                 by = c("country")
                                 )
dim(LifeExpectancyYrs_FemaleLiteracyRate_WaterSourcePrct_2011)
LifeExpectancyYrs_FemaleLiteracyRate_WaterSourcePrct_2011
skim(LifeExpectancyYrs_FemaleLiteracyRate_WaterSourcePrct_2011)
```


### save joined dataset

```{r}
save(LifeExpectancyYrs_FemaleLiteracyRate_WaterSourcePrct_2011, file = "LifeExpectancyYrs_FemaleLiteracyRate_WaterSourcePrct_2011.RData")
write_csv(LifeExpectancyYrs_FemaleLiteracyRate_WaterSourcePrct_2011, path = "LifeExpectancyYrs_FemaleLiteracyRate_WaterSourcePrct_2011.csv")
```


## All 2011 datasets


```{r}
vars_2011 <- as.data.frame(ls()) %>% filter(str_detect(`ls()`, "_11"))
vars_2011
```


```{r}
join_2011 <- Reduce(full_join,list(
  CO2emissions_11, ElectricityUsePP_11, FoodSupplykcPPD_11, IncomePP_11, LifeExpectancyYrs_11, FemaleLiteracyRate_11, population_11, WaterSourcePrct_11
  ))

dim(join_2011) # lots of missing values
join_2011 # lots of missing values

join_2011 %>% kable()
```



# Regions

```{r}
geog_temp <- read_csv("raw_data/Geography_data.csv")
geog <- clean_names(geog_temp)
dim(geog)
names(geog)
glimpse(geog)

geog
skim(geog)

geog <- geog %>% rename(country = name)
names(geog)
```


## Add regions data to 2011 variables

```{r}
dim(join_2011)
dim(geog)

join_2011_regions <- left_join(join_2011, geog,
                               by = "country")

dim(join_2011_regions)
```

  
### save joined dataset

```{r}
save(join_2011_regions, file = "Gapminder_vars_2011.RData")
write_csv(join_2011_regions, path = "Gapminder_vars_2011.csv")
```



