---
title: "Day 2 R code for Intro to R EDA workshop"
author: "J Minnier & M Niederhausen"
date: "9/17/2020"
output:
  html_document:
    number_sections: yes
    theme: lumen
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
```

* This file contains just the R code for Day 2 of the workshop.
* fyi, EDA = Exploratory Data Analysis

# Load the data

```{r}
# This chunk will only load the data if the penguins.csv file is in the same directory as this Rmd file.
penguins <- read_csv("penguins.csv")
```

# Working with data: the pipe `%>%`

```{r}
penguins %>% head(n=3)      # prounounce %>% as "then"

penguins %>% head(n=2) %>% summary()
```

```{r}
mean(penguins$body_mass_g)
median(penguins$body_mass_g)

penguins %>%
  summarize(mean(body_mass_g), 
            median(body_mass_g)) 

penguins %>%
  summarize(mean_mass = mean(body_mass_g), 
            mean_len = mean(bill_length_mm, na.rm = TRUE)) 

```

```{r}
# summary of all data as a whole
penguins %>% 
  summarize(mass_mean =mean(body_mass_g), 
            mass_sd = sd(body_mass_g),  
            mass_cv = sd(body_mass_g)/mean(body_mass_g)) 
```

```{r}
# summary by group variable
penguins %>% 
  group_by(species) %>% 
  summarize(n_per_group = n(), 
            mass_mean =mean(body_mass_g),
            mass_sd = sd(body_mass_g),
            mass_cv = sd(body_mass_g)/mean(body_mass_g))

```

## `accross()`

```{r}
penguins %>% 
  summarize(across(c(body_mass_g, bill_depth_mm), mean))

```

```{r}
penguins %>%
  summarize(across(c(bill_length_mm, bill_depth_mm), mean, na.rm=TRUE))

```

```{r}
penguins %>% 
  summarize(across(c(body_mass_g, bill_depth_mm), 
                   c(m = mean, sd = sd))) 

```

```{r}
penguins %>%
  summarize(across(where(is.character), n_distinct)
            )

```

```{r}
penguins %>%
  summarize(across(where(is.numeric), min, na.rm=TRUE))
```


## `count()`
```{r}
penguins %>% count(island)

penguins %>% count(species, island)
```


## `tabyl()`

```{r}
# default table
penguins %>% tabyl(species)

```

```{r}
# output can be treated as tibble
penguins %>% tabyl(species) %>% select(-n)

```

```{r}
penguins %>% 
  tabyl(species) %>%
  adorn_totals("row") %>% 
  adorn_pct_formatting(digits=2)  

```


```{r}
# default 2x2 table
penguins %>% 
  tabyl(species, sex) 
```


```{r}
penguins %>% tabyl(species, sex) %>% 
  adorn_percentages(denominator = "col") %>% 
  adorn_totals("row") %>% 
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns() 

```


```{r}
penguins %>% tabyl(species, island, sex)
```


```{r}
penguins %>% summarize(n_distinct(year))
```

```{r}
penguins %>% count(year)
```

```{r}
penguins %>% 
  group_by(species, sex) %>%
  summarize(median(body_mass_g))

```

```{r}
penguins %>% tabyl(island, year)
```

# Data Wrangling

## `filter()`

```{r}
penguins %>% filter(bill_length_mm > 55)


penguins %>% filter(island == "Torgersen")
penguins %>% filter(bill_length_mm/bill_depth_mm > 3)    # can do math
penguins %>% filter((body_mass_g < 3000) | (body_mass_g > 6000))

# filter on multiple variables:
penguins %>% filter(body_mass_g < 3000, bill_depth_mm < 20, sex == "female")
penguins %>% filter(body_mass_g < 3000 & bill_depth_mm < 20 & sex == "female")
penguins %>% filter(body_mass_g < 3000 | bill_depth_mm < 20 | sex == "female")

penguins %>% filter(year == 2008)      # note the use of == instead of just =
penguins %>% filter(sex == "female")

penguins %>% filter(!(species == "Adelie"))
penguins %>% filter(species %in% c("Chinstrap", "Gentoo"))

penguins %>% filter(is.na(sex))
penguins %>% filter(!is.na(sex))
```

## `select()`

```{r}
penguins %>% select(id, island, species, body_mass_g)


penguins %>% select(id:bill_length_mm)

penguins %>% select(where(is.character))
penguins %>% select(where(is.numeric))

penguins %>% select(-species)
names(penguins)
penguins %>% select(-(island:sex))

penguins %>% select(contains("bill"))
penguins %>% select(starts_with("s"))
penguins %>% select(-contains("mm"))

```

## `relocate()`

```{r}
penguins %>% relocate(year, body_mass_g)

penguins %>% relocate(species:bill_length_mm)

penguins %>% relocate(where(is.character))
penguins %>% relocate(where(is.numeric))

penguins %>% relocate(flipper_length_mm,.before = bill_length_mm)
penguins %>% relocate(species, .after = island)
penguins %>% relocate(species, .after = last_col())

```

## Save new dataset

```{r}
penguins_sub <- penguins %>% select(id:island, sex)
penguins_sub

```

## `mutate()`

```{r}
penguins <- penguins %>% 
  mutate(bill_ratio = bill_length_mm / bill_depth_mm)   
# use = (not <- or ==) to define new variable

penguins %>% select(bill_ratio, bill_length_mm, bill_depth_mm)


penguins <- penguins %>% mutate(bill_long = (bill_length_mm > 45))

penguins <- penguins %>% mutate(male = (sex == "male"))
penguins <- penguins %>% mutate(male2 = 1 * (sex == "male"))
```

## `rename()`

```{r}
# This does not save the new name
penguins %>% rename(record = id)


penguins2 <- penguins %>% 
  rename(record = id)
penguins2

```

# Figures with ggplot

## Simple plots

```{r}
ggplot(data = penguins, 
       aes(x = flipper_length_mm, 
           y = bill_length_mm)) +
  geom_point() 


ggplot(data = penguins, 
       aes(x = flipper_length_mm)) +
  geom_histogram() 
```

## Scatterplot

```{r}
ggplot(data = penguins,
       aes(x = flipper_length_mm,
           y = bill_length_mm,
           color = species)) +
  geom_point()+
  labs(
    title = "Flipper & bill length",
    subtitle = "Palmer Station LTER",
    x = "Flipper length(mm)",
    y = "Bill length(mm)") +
  scale_color_viridis_d(
    name = "Penguin species") +
  theme_bw()
```

## Histogram

```{r}
ggplot(data = penguins,
       aes(x = flipper_length_mm,
           fill = species)) +
  geom_histogram(
    alpha = 0.5,
    position = "identity") +
  labs(
    title = "Flipper length",
    x = "Flipper length(mm)",
    y = "Frequency") +
  scale_fill_viridis_d(
    name = "Penguin species") +
  theme_minimal()

```

## Boxplot + jitter

```{r}
ggplot(data = penguins,
       aes(x = species,
           y = flipper_length_mm)) +
  geom_boxplot(color="darkgrey",
               width = 0.3,
               show.legend = FALSE) +
  geom_jitter(
    aes(color = species),
    alpha = 0.5,
    show.legend = FALSE,
    position = position_jitter(
      width = 0.2, seed = 0)) +
  scale_color_manual(
    values = c("darkorange","purple",
               "cyan4")) +
  theme_minimal() +
  labs(x = "Species",
       y = "Flipper length (mm)")
```

## Bar plots

### With counts

```{r}
ggplot(data = penguins,
       aes(x = species, 
           fill = sex)) +
  geom_bar()
```

```{r}
ggplot(data = penguins,
       aes(x = species, 
           fill = sex)) +
  geom_bar(position = "dodge")

```


### With percentages

```{r}
pct_data <- penguins %>% 
  count(species, sex) %>% 
  # filter(!is.na(sex)) %>%
  group_by(species) %>% 
  mutate(pct = 100*n/sum(n))
pct_data

ggplot(data = pct_data, 
       aes(x = species, y = pct, 
           fill = sex)) +
  geom_col()+ 
  ylab("Percent")
```

```{r}
pct_data <- penguins %>% 
  count(species, sex) %>% 
  # filter(!is.na(sex)) %>%
  group_by(species) %>% 
  mutate(pct = 100*n/sum(n))
pct_data

ggplot(data = pct_data, 
       aes(x = species, y = pct, 
           fill = sex)) +
  geom_col(position = "dodge") + 
  ylab("Percent")
```


# Factor variables

```{r}
penguins <- penguins %>%
  mutate(sex_fac = factor(sex)) 
levels(penguins$sex_fac)  # factor levels are in alphanumeric order by default
penguins %>% select(sex, sex_fac) %>% summary()  # character vs. factor types
penguins %>% select(sex, sex_fac) %>% str()  # str for structure
```

## relevel 

```{r}
penguins <- penguins %>%
  mutate(species_fac = factor(species)) 

summary(penguins$species_fac)  # levels are in alphanumeric order by default

penguins <- penguins %>%
  mutate(species_fac = fct_relevel(species_fac,   
                                   c("Adelie", "Gentoo", "Chinstrap"))) 

summary(penguins$species_fac)  # levels are specified order

```

## collapse factor levels

```{r}
penguins <- penguins %>%
  mutate(species_fac2 = fct_collapse(species_fac, # collapse levels 
                                     Adelie = c("Adelie"), 
                                     Other = c("Gentoo", "Chinstrap")) 
  )

penguins %>% select(species_fac, species_fac2) %>% summary()
penguins %>% tabyl(species_fac, species_fac2) 
```


# Save data frame
```{r}
save(penguins, file = "penguins.RData")  # saving mydata within the data folder

load("penguins.RData")

write_csv(penguins, path = "penguins_new.csv")

```

