---
title: An Introduction to R and RStudio for Exploratory Data Analysis
author: "Jessica Minnier, PhD & Meike Niederhausen, PhD<br><span style = 'font-size: 80%;'>[OCTRI Biostatistics, Epidemiology, Research & Design (BERD) Workshop](https://www.ohsu.edu/octri/octri-research-forum-your-monthly-clinical-and-translational-research-event) </span>"
date: "<span style = 'font-size: 80%;'>Part 1: 2020/09/16 & Part 2: 2020/09/17 <br><br><br> slides: [bit.ly/berd_intro_part2](http://bit.ly/berd_intro_part2) <br> pdf: [bit.ly/berd_intro_part2_pdf](http://bit.ly/berd_intro_part2_pdf)</span>"
output: 
  xaringan::moon_reader:
    css: [css/xaringan-themer.css, css/my-theme.css]
    lib_dir: libs
    nature:
      highlightStyle: tomorrow #http://arm.rbind.io/slides/xaringan.html#77
      highlightLines: true
      highlightLanguage: r
      countIncrementalSlides: false
      titleSlideClass: ["left", "middle", "inverse"]
      ratio: "16:10"
    includes:
      in_header: ../header.html   
editor_options: 
  chunk_output_type: console
---

layout: true
  
<!-- <div class="my-footer"><span>bit.ly/berd_tidy</span></div>  -->

---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

library(tidyverse)
library(lubridate)
library(janitor)
library(emo)
library(here)

knitr::opts_chunk$set(
  warning=FALSE, 
  message=FALSE, 
  #fig.width=10.5, 
  #fig.height=4,
  fig.align = "center",
  rows.print=7,
  echo=TRUE,
  highlight = TRUE,
  # prompt = TRUE, # IF TRUE adds a > before each code input
  # comment = NA # PRINTS IN FRONT OF OUTPUT, default is '##' which comments out output
  comment = '##'
  )

# set ggplot theme
# theme_set(theme_bw(base_size = 24))
```

```{r xaringan-themer, include = FALSE}
# creates xaringan theme
# devtools::install_github("gadenbuie/xaringanthemer")
library(xaringanthemer)
mono_light(
  base_color =  "#3A6185", ## OHSU Marquam
  code_highlight_color = "#cbdded",
  link_color = "#38BDDE",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i","400i","700"),
  code_font_google   = NULL,
  text_font_size = "24px",
  code_font_size = "20px",
  header_h1_font_size = "45px",
  header_h2_font_size = "40px",
  header_h3_font_size = "35px",
  padding = "0em 2em 1em 2em",
  outfile = "css/xaringan-themer.css"
)
```

```{r, include=FALSE}
library(tidyverse)
library(janitor)
penguins <- read_csv(here::here("01-intro-r-eda","data","penguins.csv"))
```


---

```{r child=here::here("01-intro-r-eda","child_wrangling.Rmd")}
# Meike
```

```{r child=here::here("01-intro-r-eda","child_practice3.Rmd")}
# Meike - mutate/filter etc
```

```{r child=here::here("01-intro-r-eda","child_ggplot.Rmd")}
```

```{r child=here::here("01-intro-r-eda","child_factors.Rmd")}
```

---

# Save data frame

* Save __.RData__ file: the standard R format, which is recommended if saving data for future use in R
```{r, eval=FALSE}
save(mydata, file = "data/mydata.RData")  # saving mydata within the data folder
```

You can load .RData files using the load() command:

```{r, eval=FALSE}
load("data/mydata.RData")
```
  

<br>

* Save __csv__ file: comma-separated values

```{r, eval=FALSE}
write.csv(mydata, file = "data/mydata.csv", col.names = TRUE, row.names = FALSE)
```

---

class: center, middle, inverse

# More commands to filter rows

<!-- TO-DO: rename? -->

---

# Remove rows with missing data

`drop_na()` (or `na.omit()`) removes *all* rows with *any* missing (`NA`) values in *any* column

```{r}
mydata %>% drop_na()
```

For more on dealing with missing data, see [Tidyverse part 2 workshop](https://jminnier-berd-r-courses.netlify.app/02-data-wrangling-tidyverse/02_data_wrangling_slides_part2.html#35)

---

# Remove rows with duplicated data

`distinct()` removes rows that are duplicates of other rows

```{r}
data_dups <- tibble(
  name = c("Ana","Bob","Cara", "Ana"),
  race = c("Hispanic","Other", "White", "Hispanic")
)
```

.pull-left[
```{r}
data_dups
```
]
.pull-right[
```{r}
data_dups %>% distinct()
```
]

---

# Order rows: `arrange()`

Use `arrange()` to order the rows by the values in specified columns

```{r}
mydata %>% arrange(bmi, weight_kg) %>% head(n=3)

mydata %>% arrange(desc(bmi), weight_kg) %>% head(n=3)
```



---

# Combine `summarize()` with other tidy functions

```{r}
mydata %>% 
  group_by(grade) %>% 
  summarize(n_per_group = n(), 
            bmi_mean =mean(bmi),
            bmi_sd = sd(bmi, na.rm = TRUE)) %>%
  mutate(bmi_cv = bmi_sd/bmi_mean) %>% #<<
  filter(bmi_cv > .2)  #<<
```




---

```{r child=here::here("01-intro-r-eda","child_closing.Rmd")}
```

