---
title: Practice Solutions to <br> Intro to R and Rstudio for EDA - Part 2
author: Jessica Minnier, PhD & Meike Niederhausen, PhD
date: 'OCTRI Biostatistics, Epidemiology, Research & Design (BERD) Workshop <br><br> 2020/09/17'
output: 
  xaringan::moon_reader:
    css: [css/xaringan-themer-solns.css, css/my-theme.css]
    lib_dir: libs
    nature:
      highlightStyle: tomorrow #http://arm.rbind.io/slides/xaringan.html#77
      highlightLines: true
      highlightLanguage: r
      countIncrementalSlides: false
      titleSlideClass: ["left", "middle", "inverse"]
      ratio: "16:9"
    includes:
      in_header: ../header.html   
editor_options: 
  chunk_output_type: console
---

layout: true
  
<!-- <div class="my-footer"><span>bit.ly/berd_tidy</span></div>  -->

---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning=FALSE, 
  message=FALSE, 
  #fig.width=10.5, 
  #fig.height=4,
  fig.align = "center",
  rows.print=7,
  echo=TRUE,
  highlight = TRUE,
  prompt = FALSE, # IF TRUE adds a > before each code input
  comment = NA # PRINTS IN FRONT OF OUTPUT, default is '##' which comments out output
  #comment=NA
  )

# set ggplot theme
#theme_set(theme_bw(base_size = 24))
```



```{r xaringan-themer, include = FALSE}
# creates xaringan theme
# devtools::install_github("gadenbuie/xaringanthemer")
library(xaringanthemer)
mono_light(
  base_color =  "#518c66", ## OHSU Marquam
  code_highlight_color = "#c0e8f5",
  link_color = "#38BDDE",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i","400i","700"),
  code_font_google   = NULL,
  text_font_size = "24px",
  code_font_size = "20px",
  header_h1_font_size = "45px",
  header_h2_font_size = "40px",
  header_h3_font_size = "35px",
  padding = "0em 2em 1em 2em",
  outfile = "css/xaringan-themer-solns.css"
)
```


```{r, include=FALSE}
library(tidyverse)
library(janitor)
penguins <- read_csv(here::here("01-intro-r-eda","data","penguins.csv"))
mydata <- read_csv(here::here("01-intro-r-eda","data","yrbss.csv"))

```


# Practice 3

1. Continue adding code chunks to your Rmd (or, start a new one! But remember to load the libraries and data at the top.)

1. How many different years are in the data? (Hint: use `tabyl()` or `n_distinct()`)

1. Count the number of penguins measured each year.

1. Calculate the median body mass by each species and sex subgroup. Use `summarize()` and `group_by()` to do this.

1. Create a 2x2 table of number of penguins measured in each year by each island.

---

# Practice 3 Answers

### 2. How many different years are in the data? (Hint: use `tabyl()` or `n_distinct()`)

.pull-left[
Option 1:
```{r}
penguins %>% 
  summarize(n_distinct(year))
```
]
.pull-right[
Option 2:
```{r}
penguins %>% tabyl(year)
penguins %>% tabyl(year) %>% nrow
```
]

---

### 3. Count the number of penguins measured each year.

.pull-left[
Option 1:
```{r}
penguins %>% count(year)
```
]
.pull-right[
Option 2:
```{r}
penguins %>% tabyl(year)
```
]

---

### 4. Calculate the median body mass by each species and sex subgroup. Use `summarize()` and `group_by()` to do this.

```{r}
penguins %>% 
  group_by(species, sex) %>%
  summarize(median(body_mass_g))
```

---

### 5. Create a 2x2 table of number of penguins measured in each year by each island.

```{r}
penguins %>% tabyl(island, year)
```

---

# Practice 4 (old practice 2)

1. Create a new script and save it as `Practice2.R`

1. Create data frames for males and females separately.

1. Do males and females have similar BMIs? Weights? Compares means, standard deviations, range, and boxplots.

1. Plot BMI vs. weight for each gender separately. Do they have similar relationships?

1. Are males or females more likely to be bullied in the past 12 months? Calculate the percentage bullied for each gender.


---

# Practice 4 Answers

__#2__ Create data frames for males and females separately.

```{r}
boys <- mydata[mydata$sex == "Male", ]
dim(boys)
girls <- mydata[mydata$sex == "Female", ]
dim(girls)
```

Check number of boys & girls:
```{r}
summary(mydata$sex)
```


---

__#3__ Do males and females have similar BMIs? Weights? Compares means, standard deviations, range, and boxplots.

.pull-left-60[
```{r}
summary(boys$bmi); sd(boys$bmi)
summary(girls$bmi); sd(girls$bmi)
```

]
.pull-right-40[

```{r fig.height=5}
boxplot(mydata$bmi ~ mydata$sex)
```
]

---


__#4__ Plot BMI vs. weight for each gender separately. Do they have similar relationships?

.pull-left[
```{r fig.height=5, fig.width=4}
plot(boys$bmi, boys$weight)
```

]

.pull-right[

```{r fig.height=5, fig.width=4}
plot(girls$bmi, girls$weight)
```

]



---

__#5__ Are males or females more likely to be bullied in the past 12 months? Calculate the percentage bullied for each gender.

.pull-left[
```{r}
bullied_boys <- 
  boys[boys$bullied_past_12mo == TRUE,]
nrow(bullied_boys)
bullied_boys_prct <- 
  nrow(bullied_boys) / nrow(boys) * 100
bullied_boys_prct

# alternative
mean(boys$bullied_past_12mo, na.rm=TRUE)
```
]
.pull-right[
```{r}
# Apply the same method for girls:
bullied_girls <- 
  girls[girls$bullied_past_12mo == TRUE,]
nrow(bullied_girls)
bullied_girls_prct <- 
  nrow(bullied_girls) / nrow(girls) * 100
bullied_girls_prct

# alternative. Answers don't match. Why???
mean(girls$bullied_past_12mo, na.rm=TRUE)
```


]


---

__#5__ cont'd

On the previous slide we saw that our two methods for calculating the percentage of girls that were bullied in the past 12 months did not match. What went wrong?


```{r}
nrow(bullied_girls)
girls$bullied_past_12mo
```


To get the number of girls that were bullied we need to make sure the missing values (NA) are not included.

---

__#5__ cont'd - working with NA's

```{r}
# values of bullied_past_12mo
girls$bullied_past_12mo

# which are missing (logical)
is.na(girls$bullied_past_12mo)

# which are NOT missing (logical)
!is.na(girls$bullied_past_12mo)
```

---

__#5__ cont'd - fix girls' code

Exclude the missing values from the `bullied_girls`:

```{r}
girls2 <- girls[!is.na(girls$bullied_past_12mo),]
nrow(girls2)

bullied_girls2 <- girls2[girls2$bullied_past_12mo == TRUE,]
nrow(bullied_girls2)

# from girls dataset, total number bullied
sum(girls$bullied_past_12mo, na.rm = TRUE)
```

---

__#5__ cont'd - Calculate percentage girls bullied

```{r}
bullied_girls_prct2 <- nrow(bullied_girls2) / nrow(girls2) * 100
bullied_girls_prct2

# Compare to alternative
mean(girls$bullied_past_12mo, na.rm=TRUE)
```

---

# Practice 5

1. Continue adding code chunks to your Rmd (or, start a new one! But remember to load the libraries and data at the top.)

1. Make a scatter plot of bill depth vs bill length.

1. Add `+ geom_smooth(method="lm")` to the plot. What is this saying about the association between bill depth and length?

1. Now add `color = species` to the aesthetic `aes()`. Keep `geom_smooth`. How do the associations look now?


---

### 2. Make a scatter plot of bill depth vs bill length.

.pull-left[
```{r ex5_2, eval = FALSE}
ggplot(data = penguins, 
       aes(x = bill_length_mm, 
           y = bill_depth_mm)) + 
  geom_point()
```
]
.pull-right[
```{r ex5_2_out, ref.label="ex5_2", echo=FALSE, fig.keep = "first"}
```
]

---

### 3. Add `+ geom_smooth(method="lm")` to the plot. What is this saying about the association between bill depth and length?

.pull-left[
It looks like as bill length increases, bill depth decreases. This is a negative association. But there is a wide variation and a lot of noise.

```{r ex5_3, eval = FALSE}
ggplot(data = penguins, 
       aes(x = bill_length_mm, 
           y = bill_depth_mm)) + 
  geom_point()+
  geom_smooth(method="lm")
```
]
.pull-right[
```{r ex5_3_out, ref.label="ex5_3", echo=FALSE, fig.keep = "first"}
```
]

---

### 4. Now add `color = species` to the aesthetic `aes()`. Keep `geom_smooth`. How do the associations look now?

.pull-left[
The association reverses, when we look inside species. As bill length increases, bill depth increases. This is an example of [Simpson's paradox!](https://en.wikipedia.org/wiki/Simpson%27s_paradox#:~:text=Simpson's%20paradox%2C%20which%20goes%20by,when%20these%20groups%20are%20combined.)

```{r ex5_4, eval = FALSE}
ggplot(data = penguins, 
       aes(x = bill_length_mm, 
           y = bill_depth_mm,
           color = species)) + #<<
  geom_point()+
  geom_smooth(method="lm")
```
]
.pull-right[
```{r ex5_4_out, ref.label="ex5_4", echo=FALSE, fig.keep = "first"}
```
]

---

### Here's a prettier version:

.pull-left[
```{r ex4_p, eval = FALSE}
ggplot(data = penguins, 
       aes(x = bill_length_mm, 
           y = bill_depth_mm,
           color = species)) + 
  geom_point()+
  geom_smooth(method="lm")+
  labs( 
    title = "Flipper vs bill length",
    subtitle = "Palmer Station LTER",
    x = "Flipper length(mm)", 
    y = "Bill length(mm)") + 
  scale_color_viridis_d( 
    name = "Species") + 
  theme(legend.position = "bottom")+
  theme_bw()
```
]
.pull-right[
```{r ex4_p_out, ref.label="ex4_p", echo=FALSE, fig.keep = "first", fig.width = 8, fig.height = 8}
```
]







```{r, include=FALSE}
#knitr::purl("01-getting-started-v2/01_getting_started_Practice_Answers.Rmd")
pagedown::chrome_print(here::here("01-intro-r-eda","01_intro_r_eda_Practice_Answers_part2.html"))

```

