


---
class: inverse, center, middle

# Working with the data

---

# The $

Suppose we want to single out the column of bill length values.

.pull-left[
* How did we previously learn to do this?
```{r}
penguins[, 4]
```
]
.pull-right[
The problem with this method, is that we need to know the column number which can change as we make changes to the data set. 

* Use the `$` instead: `DatSetName$VariableName`

```{r}
penguins$bill_length_mm
```
]
---

# Basic plots of numeric data: Histogram

```{r fig.height=2.5, fig.width=7}
hist(penguins$bill_length_mm)
```

With extra features: 

```{r fig.height=2.5, fig.width=7}
hist(penguins$bill_length_mm, xlab = "Length (mm)", main="Penguin bills")
```

---

# Basic plots of numeric data: Boxplot

.pull-left[
```{r fig.height=5, fig.width=5}
boxplot(penguins$bill_length_mm)
```

]

--

.pull-right[
```{r fig.height=5, fig.width=5}
boxplot(penguins$bill_length_mm ~ 
          penguins$sex, 
  horizontal = TRUE, 
  xlab = "Length (mm)", ylab = "Sex",
  main = "Penguin bills by sex")
```
]

---

# Basic plots of numeric data: Scatterplot

.pull-left[
```{r fig.height=5, fig.width=5}
plot(penguins$flipper_length_mm, 
     penguins$bill_length_mm)
```
]

.pull-right[
```{r fig.height=5, fig.width=5}
plot(penguins$flipper_length_mm, 
     penguins$bill_length_mm, 
   xlab = "Flipper", ylab = "Bill", 
   main = "Bill vs. flipper length")
```
]

---

# Summary stats of numeric data (1/3)

* Standard R `summary` command

```{r}
summary(penguins$flipper_length_mm)
```

* Mean and standard deviation

```{r}
mean(penguins$flipper_length_mm)
sd(penguins$flipper_length_mm)
```

---

# Summary stats of numeric data (2/3)
<!-- QQ: Why is (2/2) being cut off? It's not cut off for (1/2).-->
* Min, max, & median

.pull-left[
```{r}
min(penguins$flipper_length_mm)
max(penguins$flipper_length_mm)

```
]

.pull-right[
```{r}
median(penguins$flipper_length_mm)
```
]

* Quantiles

```{r}
quantile(penguins$flipper_length_mm, prob=c(0, .25, .5, .75, 1))  
```



---

# Summary stats of numeric data (3/3)

.pull-left-60[
* Find the mean bill length

```{r}
mean(penguins$bill_length_mm)
```

*Why did we get `NA` for the mean?*

]

.pull-right-40[

Since there are missing values (`NA`), we need to tell R to remove them from the data when calculating the mean.

```{r}
mean(penguins$bill_length_mm, 
     na.rm = TRUE) #<<
```
]


```{r}
summary(penguins$bill_length_mm)
```


