

---



class: center, middle, inverse

# More commands to filter rows

<!-- TO-DO: rename? -->

---

# Remove rows with missing data

`drop_na()` (or `na.omit()`) removes *all* rows with *any* missing (`NA`) values in *any* column

```{r}
mydata %>% drop_na()
```

For more on dealing with missing data, see [Tidyverse part 2 workshop](https://jminnier-berd-r-courses.netlify.app/02-data-wrangling-tidyverse/02_data_wrangling_slides_part2.html#35)

---

# Remove rows with duplicated data

`distinct()` removes rows that are duplicates of other rows

```{r}
data_dups <- tibble(
  name = c("Ana","Bob","Cara", "Ana"),
  race = c("Hispanic","Other", "White", "Hispanic")
)
```

.pull-left[
```{r}
data_dups
```
]
.pull-right[
```{r}
data_dups %>% distinct()
```
]

---

# Order rows: `arrange()`

Use `arrange()` to order the rows by the values in specified columns

```{r}
mydata %>% arrange(bmi, weight_kg) %>% head(n=3)

mydata %>% arrange(desc(bmi), weight_kg) %>% head(n=3)
```



---

# Combine `summarize()` with other tidy functions

```{r}
mydata %>% 
  group_by(grade) %>% 
  summarize(n_per_group = n(), 
            bmi_mean =mean(bmi),
            bmi_sd = sd(bmi, na.rm = TRUE)) %>%
  mutate(bmi_cv = bmi_sd/bmi_mean) %>% #<<
  filter(bmi_cv > .2)  #<<
```


