


---

class: inverse, center, middle

# Working with the data

---

# The $

Suppose we want to single out the column of BMI values.

.pull-left[
* How did we previously learn to do this?
```{r}
penguins[, 4]
```
]
.pull-right[
The problem with this method, is that we need to know the column number which can change as we make changes to the data set. 

* Use the `$` instead: `DatSetName$VariableName`

```{r}
penguins$bill_length_mm
```
]
---

# Basic plots of numeric data: Histogram

```{r fig.height=2.5, fig.width=7}
hist(mydata$bmi)
```

With extra features: 

```{r fig.height=2.5, fig.width=7}
hist(mydata$bmi, xlab = "BMI", main="BMIs of students")
```

---

# Basic plots of numeric data: Boxplot

.pull-left[
```{r fig.height=5, fig.width=5}
boxplot(mydata$bmi)
```

]

--

.pull-right[
```{r fig.height=5, fig.width=5}
boxplot(mydata$bmi ~ mydata$sex, 
  horizontal = TRUE, 
  xlab = "BMI", ylab = "sex",
  main = "BMIs of students by sex")
```
]

---

# Basic plots of numeric data: Scatterplot

.pull-left[
```{r fig.height=5, fig.width=5}
plot(mydata$weight_kg, mydata$bmi)
```
]

.pull-right[
```{r fig.height=5, fig.width=5}
plot(mydata$weight_kg, mydata$bmi, 
   xlab = "weight (kg)", ylab = "BMI", 
   main = "BMI vs. Weight")
```
]

---

# Summary stats of numeric data (1/2)

* Standard R `summary` command

```{r}
summary(mydata$bmi)
```

* Mean and standard deviation

```{r}
mean(mydata$bmi)
sd(mydata$bmi)
```

---

# Summary stats of numeric data (2/2)
<!-- QQ: Why is (2/2) being cut off? It's not cut off for (1/2).-->
* Min, max, & median

.pull-left[
```{r}
min(mydata$bmi)
max(mydata$bmi)

```
]

.pull-right[
```{r}
median(mydata$bmi)
```
]

* Quantiles

```{r}
quantile(mydata$bmi, prob=c(0, .25, .5, .75, 1))  
```


<!-- # Add height column to data frame -->

<!-- Since $\textrm{BMI} = \frac{kg}{m^2}$, we have -->
<!-- $\textrm{height}(m) = \sqrt{\frac{\textrm{weight}(kg)}{\textrm{BMI}}}$ -->


<!-- mydata$height_m <- sqrt( mydata$weight_kg / mydata$bmi ) -->
<!-- mydata$height_m -->

